The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-12-01 article_final/main.tex
Final version done

%\documentclass[wcp,gray]{jmlr} % test grayscale version
 %\documentclass[wcp]{jmlr}% former name JMLR W\&CP
\documentclass[pmlr]{jmlr}% new name PMLR (Proceedings of Machine Learning)
\input{math_commands.tex}
\RequirePackage{graphicx}
 % The following packages will be automatically loaded:
 % amsmath, amssymb, natbib, graphicx, url, algorithm2e
 \usepackage{booktabs}
 %\usepackage{rotating}% for sideways figures and tables
\usepackage{longtable}% for long tables
\usepackage{xcolor}% required by colortbl
\usepackage{colortbl}% for colored table rules (required by arydshln)
\usepackage{arydshln}% for dashed lines in tables

@@

\emph{In deriving Muon's update step, why should one constrain by the spectral or any other operator norm? How would alternative norms affect performance and computational cost?}

In this article, we tackle this question by showing that there are many viable non-operator norms. We leverage the family of norms dual to Ky Fan $k$-norms to derive a new family of \emph{Fanions}, algorithms with low-rank updates. This approach theoretically explains the backbone of Dion's update \citep{ahn2025dion} and generalizes the memory-motivated application of the nuclear norm to Sharpness-Aware Minimization \citep{pethick2025sam}. As was done with Muon, we develop an effective procedure for computing Fanions' updates. The Lanczos algorithm, which is described and compared with its competitors in \sectionref{sec:matrix-side}, is the most operation-efficient algorithm, though it currently lacks an effective GPU- and PyTorch-friendly implementation.

@@

After that in \sectionref{sec:experiments} compare the performance of these algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT and GPT-2 Medium on FineWeb dataset \citep{modded_nanogpt_2024}
	\item Fine-tuning NanoGPT on Tiny Stories dataset \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

@@

Most notably, our experiments on real-world tasks demonstrate that the choice of underlying matrix norm is remarkably flexible. On CIFAR-10 airbench, properly-tuned F-Muon and S-Muon achieve $94.02 \pm 0.13\%$ and $94.03 \pm 0.13\%$ accuracy, matching Muon's $94.01 \pm 0.13\%$ performance. On pre-training of NanoGPT F-Muon achieves 3.281 cross-entropy loss, while fully-tuned Muon achieves 3.279; of GPT-2 Medium - F-Muon 2.9215 and Muon 2.9198. Finally, S-Muon matches Muon on fine-tuning NanoGPT on Tiny Stories, while F-Muon is far more resistant to learning rate choice than Muon. These results show that Muon-like algorithms can maintain competitive performance even when the underlying norm constraint is significantly modified, providing an affirmative answer to the central question posed above. Moreover, the tools from \sectionref{sec:conic_combo} give researchers considerable flexibility in designing algorithms that need not be direct modifications of Muon.

@@

Such problems can be solved with an iterative algorithm based on the Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
	\mathrm{LMO}(\mM^t) \in \argmin_{\mD \in \cS} \langle\mM^t, \mD\rangle\,,
\end{equation}
where $\mM^t$ is the effective update direction and $\cS \subset \Rmn$ is a constraint set. The algorithm proceeds as:
\begin{equation}\label{eq:simple_update}
  \begin{aligned}
  \mB^t &= \beta \mB^{t-1} + (1 - \beta) \nabla F(\mX^t) \quad \text{(momentum buffer)},\\
  \mM^t &= \begin{cases}
  \nabla F(\mX^t) & \text{(no momentum)},\\
  \mB^t & \text{(heavy ball)},\\
  \nabla F(\mX^t) + \beta \mB^t & \text{(approximate Nesterov)},
  \end{cases}\\
  \mX^{t + 1} &= \mX^t + \gamma_t\mathrm{LMO}\left(\mM^t\right)\,,
  \end{aligned}
\end{equation}
where $\beta \in [0, 1)$ is the momentum coefficient. Throughout our experiments, we employ approximate Nesterov momentum, which matches the implementation in Muon and PyTorch SGD (\texttt{nesterov=True}). It uses the current gradient rather than the true lookahead gradient, trading theoretical guarantees for computational efficiency.

@@

Using this formula, it is easy to compute updates for algorithms induced by various norms $\norm{\cdot}$:

@@

\paragraph{Schatten norms}
\citep{cesista2025schattenp} considered Schatten-$p$ norms:
\[\norm{\mM^t}_{S_p} = \left(\sum_{i=1}^{\min(m, n)}\sigma_i^p\right)^{1/p}\,,
\]
which produce the update
\[
\mX^{t+1} = \mX^t - \gamma_t \mU \frac{\diag\left(\sigma_1^{q-1},\dots, \sigma_{\min(m,n)}^{q-1}\right)}{\left(\sum_{i=1}^{\min(m,n)}{\sigma_i^q}\right)^{\frac{q-1}{q}}}\mV^\top
\]
where $p$ and $q$ satisfy $p^{-1} + q^{-1} = 1$. This formula recovers Neon when $p\rightarrow1$ (provided $\sigma_1 > \sigma_2$, which holds on real data), NSGD when $p=2$, and Muon when $p\rightarrow\infty$.

However, Schatten norms do not fill the rank gap: when $p > 1$, the update has full rank, while when $p=1$, it has rank one. Moreover, computing the update for $p \neq 1, 2, \infty$ appears to require knowing all $\sigma_i$, making the problem as hard as computing the full SVD.

@@

To derive the update formula for arbitrary $k$, we use the expression for the norm dual to the Ky Fan $k$-norm (see, for instance \citep{bhatia2013matrix}, p. 96):
\[
    \normkfk{\cdot}^\dagger = \max\left\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\right\}\,.
\]
According to \equationref{eq:our_update}, we need to find $\mD$ such that $\normkfk{\mD} = 1$ and
\[
\<\mM^t, \mD> = \max\left\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i,\ \sigma_1\right\}\,.
\]
The Neon update $\mD = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\mD = \frac{1}{k}\mU\mV^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
\[
\mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
\]
depending on which of those two expressions achieves smaller residual on the linear function $L(\mX) := F(\mX^t) + \<\mM^t, \mX - \mX^t>$. The Ky Fan norms thus fail to close the rank gap: the resulting update is either rank-one or full-rank.

@@

In \sectionref{sec:matrix-side}, we will discuss how to efficiently compute Fanion updates using the Lanczos algorithm.

@@

Simply applying norms dual to Ky Fan $k$-norms, however, does not provide sufficient algorithmic diversity for our purposes. We now consider linear combinations of LMO-based algorithms and show that these combinations are themselves LMO-algorithms.

@@

\begin{corollary}\label{corollary:lin_comb_alg}
	Let there be a finite family of LMO based algorithms indexed by $i = 1,\dots,n$, where the $\mX^{t + 1}$ update in the $i$-th algorithm is defined by
	\[
	\mX^{t + 1} - \mX^{t} = \gamma_t \mathrm{LMO}_{i}(\mM^{t})\,,
	\]
	and $\mathrm{LMO}_i$ corresponds to the unit ball of norm $\norm{\cdot}_i$. For arbitrary non-negative $\alpha_1,\dots,\alpha_n$, the algorithm with the update given by
	\[
	\mX^{t + 1} - \mX^{t} = \gamma_t \sum_{i = 1}^n \alpha_i \mathrm{LMO}_{\norm{\cdot}_{(i)}}(\mM^t)
	\]
	is an LMO-algorithm itself, with LMO corresponding to the unit ball of the norm $\norm{\cdot}$ dual to the norm given by $\sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}^\dagger$.
\end{corollary}

@@

The extremal members of this family are \emph{F-Neon} (with $k = 1$) and \emph{F-Muon} (with $k = \min\{m, n\}$). Additional information and visualizations for the F-Muon generating $\normfstar{\cdot}^\dagger = \norm{\cdot}_{\mathrm{F-KF-1}}^\dagger$ and the F-Neon generating $\normftwo{\cdot}^\dagger = \norm{\cdot}_{\mathrm{F-KF-\min\{m, n\}}}^\dagger$ norms appear in the appendix (see \equationref{eq:normfstardual}, \equationref{eq:normftwodual}, \figureref{fig:fduals}).

@@

The extremal members of this family are \emph{S-Neon} (with $k = 1$) and \emph{S-Muon} (with $k = \min\{m, n\}$).

\section{Computing the Updates}\label{sec:matrix-side}
We employ the thick-restart Lanczos method for the symmetric eigenvalue problem (TRLan) to compute the low-rank updates of Fanions. We apply TRLan to either $\mM^{t\top}\mM^t$ or $\mM^t \mM^{t\top}$, selecting whichever matrix is smaller. We use the CuPy implementation of \texttt{cupy.sparse.linalg.svds} \citep{cupy_svds_ref}, which internally relies on TRLan \citep{simonz1998thick}.

@@

\begin{table}[ht]
\floatconts
  {tbl:matrix_methods}
  {\caption{Comparison of methods for computing rank-$k$ updates on dense random matrices (CPU, double precision). Lower is better in all columns.}}
  {%
    \begin{tabular}{|c|c|c|c|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Matrix sizes & $k$  & Method           & Time (s) & Matvecs & Iterations & $err_1$         & $err_2$       \\
      \arrayrulecolor{black}\hline
      $500 \times 500$      & 5  & Power Iterations           & 0.062   & 2005    & 200        & $9.2 \cdot 10^{-3}$ & $9.1 \cdot 10^{-3}$\\
      $500 \times 500$      & 5  & RSVD             & 0.017   & 1170    & 38         & $9.8 \cdot 10^{-3}$ & $9.6 \cdot 10^{-3}$\\
      $500 \times 500$      & 5  & TRLan            & 0.018   & 131     & 65         & $9.6 \cdot 10^{-5}$ & $9.4 \cdot 10^{-5}$\\
      \hdashline
      $500 \times 500$      & 50 & Power Iterations           & 0.44    & 43750   & 437      & $9.9 \cdot 10^{-3}$ & $9.0 \cdot 10^{-3}$\\
      $500 \times 500$      & 50 & RSVD             & 0.61    & 6120    & 50      & $9.9 \cdot 10^{-3}$ & $9.1 \cdot 10^{-3}$\\
      $500 \times 500$      & 50 & TRLan            & 0.16    & 462     & 231      & $3.3 \cdot 10^{-7}$ & $3.0 \cdot 10^{-7}$\\
      \hdashline
      $5000 \times 5000$    & 5  & Power Iterations           & 9.6     & 9065    & 906      & $8.6 \cdot 10^{-3}$ & $8.6 \cdot 10^{-3}$\\
      $5000 \times 5000$    & 5  & RSVD             & 2.1     & 5640    & 187      & $9.7 \cdot 10^{-3}$ & $9.7 \cdot 10^{-3}$\\
      $5000 \times 5000$    & 5  & TRLan            & 0.70    & 205     & 102      & $7.7 \cdot 10^{-3}$ & $7.7 \cdot 10^{-3}$\\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

\begin{table}[ht]
\floatconts
  {tbl:newton_schulz_reference}
  {\caption{Newton-Schulz iterations performance on random dense matrices (for reference).}}
  {%
    \begin{tabular}{|c|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Matrix size    & Time (s) & Matvecs & Iterations & $err_1$ \\
      \arrayrulecolor{black}\hline
      $500 \times 500$     & 0.041    & 27 000  & 27         & $4.8 \cdot 10^{-3}$  \\
      $5000 \times 5000$   & 26.4     & 290 000 & 29         & $6.5 \cdot 10^{-3}$  \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

@@

We begin by evaluating F-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ represent typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, and $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed on the interval $(0, 1)$. We set $\mS=0$ and initialize $\mX^0$ with entries drawn independently from $\cN(0, 0.01)$.

@@

\begin{table}[htbp]
\floatconts
  {tbl:lls_lrs}
  {\caption{Tuned learning rates and momentum coefficients in the linear least squares experiment.}}
  {%
    \begin{tabular}{|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Algorithm & \texttt{lr}  & \texttt{momentum}  & Iterations to 0.001 loss \\
      \arrayrulecolor{black}\hline
      Muon & 0.007 & 0.5 & 1060 \\
      \hdashline
      NSGD & 0.08 & 0.95 & 1020 \\
      F-Muon & 0.015 & 0.7 & 910 \\
      \hdashline
      SignSGD & 0.016 $\times$ 0.01 & 0.95 & 2650 \\
      S-Muon & 0.011 & 0.9 & 890 \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

@@

\begin{figure}[htbp]
\floatconts
  {fig:lls}
  {\caption{Linear least squares problem for a 500x500 matrix.}}
  {%
    \subfigure[The loss]{\label{fig:lls_loss}%
      \includegraphics[width=0.8\linewidth]{figs/simple_lls/loss_long_run_16_increase.pdf}}%
    \vspace{1em}
    \subfigure[The Frobenius norm of the gradient]{\label{fig:lls_fro_grad_norm}%
      \includegraphics[width=0.8\linewidth]{figs/simple_lls/gradient_frobenius_norm_vs_iteration_long_run.pdf}}
  }
\end{figure}

Both F-Muon and S-Muon converge faster to lower loss values and achieve lower Frobenius norms of the gradient than NSGD, Muon, or SignSGD.

@@

\begin{table}[htbp]
\floatconts
  {tbl:mat_vs_vec_lmo}
  {\caption{LMO optimizers in Schatten $S_p$ norms and in $l_p$ norms. $g$ is the gradient. When it is a matrix, $g = \mU \mSigma \mV^\top$.}}
  {%
    \begingroup
    \def\arraystretch{1.2}
    \resizebox{\linewidth}{!}{%
      \begin{tabular}{|l|c|c|c|}
        \arrayrulecolor{black}\hline
        Algorithm                             & LMO constraint set $\mathcal{D}$ & LMO                                                         & Reference                            \\
        \arrayrulecolor{black}\hline
        \arrayrulecolor{black}\hline
        Normalized SGD                     & $l_2$-ball, $S_2$-ball          & $-\eta \tfrac{g}{\norm{g}_2} = -\eta \tfrac{g}{\normf{g}}$ & \citep{hazan2015beyond}              \\
        Momentum Normalized SGD            & Ball in $l_2$, or Ball in $S_2$ & $-\eta \tfrac{g}{\norm{g}_2} = -\eta\tfrac{g}{\normf{g}}$  & \citep{cutkosky2020momentum}         \\
        \arrayrulecolor{black}\hline
        SignSGD                            & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 1]{bernstein2018signsgd} \\
        Signum                             & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 3]{bernstein2018signsgd} \\
        \hdashline
        Muon                               & Ball in Spectral $S_\infty$     & $-\eta \mU\mV^\top$                                             & \citep{jordan2024muon}               \\
        \arrayrulecolor{black}\hline
        Gauss-Southwell Coordinate Descent & Ball in $l_1$                   & $-\eta \sum_{i \in \argmax|g_i^t|} \sign(g_i^t)e_i$                       & \citep[p. 19]{shi2016primer}         \\
        \hdashline
        Neon                               & Ball in Nuclear $S_1$           & $-\eta \vu_1 \vv_1^\top$                                        & This work                            \\
        \arrayrulecolor{black}\hline
      \end{tabular}%
    }%
    \endgroup
  }
\end{table}

@@

A common limitation of these analyses is their focus on convergence measured by the gradient norm. As we showed in our CIFAR experiments \sectionref{sec:cifar_underhood}, the gradient norm may decrease by only a factor of ten when the accuracy reaches 100\%.

@@

\subsection{The nuclear norm in the LMO}

@@

\subsection{Orthogonal research}

@@

\section{Visualization of different matrix norms}
\subsection{Duals to F* and F2 norms}

@@

\subsection{The Ky Fan norm and its dual}

@@

\section{More details for Linear Least Squares}
\label{sec:lls_plots_section}
The spectral and nuclear norms of the gradients over iterations, as well as the loss over time, are shown in \figureref{fig:app_lls}. The poor performance of Fanions and F-Fanions in terms of speed for large $k$ is likely caused by the inefficient implementation of TRLan.

>>>>

%\documentclass[wcp,gray]{jmlr} % test grayscale version
 %\documentclass[wcp]{jmlr}% former name JMLR W\&CP
\documentclass[pmlr]{jmlr}% new name PMLR (Proceedings of Machine Learning)
\input{math_commands.tex}
\RequirePackage{graphicx}
 % The following packages will be automatically loaded:
 % amsmath, amssymb, natbib, graphicx, url, algorithm2e
 \usepackage{booktabs}
\AtBeginEnvironment{table}{\setlength{\extrarowheight}{2pt}}
\AtEndEnvironment{table}{\setlength{\extrarowheight}{1pt}}
 %\usepackage{rotating}% for sideways figures and tables
\usepackage{longtable}% for long tables
\usepackage{xcolor}% required by colortbl
\usepackage{colortbl}% for colored table rules (required by arydshln)
\usepackage{arydshln}% for dashed lines in tables

@@

\emph{Can the employment of a norm other than the operator norm in Muon-like algorithms yield comparable or superior empirical results?}

In this article, we tackle this question by showing that there are many viable non-operator norms. We leverage the family of norms dual to Ky Fan $k$-norms to derive a new family of \emph{Fanions}, algorithms with low-rank updates. This approach theoretically explains the backbone of Dion's update \citep{ahn2025dion} and generalizes the memory-motivated application of the nuclear norm to Sharpness-Aware Minimization \citep{pethick2025sam}. As was done with Muon, we develop an effective procedure for computing Fanions' updates using the Lanczos algorithm (see \sectionref{sec:matrix-side}).

@@

In \sectionref{sec:experiments}, we compare the performance of these algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT and GPT-2 Medium on FineWeb dataset \citep{modded_nanogpt_2024}
	\item Fine-tuning NanoGPT on Tiny Stories dataset \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

@@

Most notably, our experiments on real-world tasks demonstrate that the choice of underlying matrix norm is remarkably flexible. On CIFAR-10 airbench, properly-tuned F-Muon and S-Muon achieve $94.02 \pm 0.13\%$ and $94.03 \pm 0.13\%$ accuracy, matching Muon's $94.01 \pm 0.13\%$ performance. After 1750 iterations on NanoGPT training, F-Muon achieves 3.281 cross-entropy loss, while Muon achieves 3.279. Similarly, after 5960 iterations of GPT-2 Medium training we get 2.9215 with F-Muon and 2.9198 with Muon. Finally, S-Muon matches Muon on fine-tuning NanoGPT on Tiny Stories, while F-Muon is far more resistant to learning rate choice than Muon. These results show that Muon-like algorithms can maintain competitive performance even when the underlying norm constraint is significantly modified, providing an affirmative answer to the central question posed above. Moreover, the tools from \sectionref{sec:conic_combo} give researchers considerable flexibility in designing algorithms that need not be direct modifications of Muon.

@@

Such problems can be solved with an iterative algorithm based on the Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
	\mathrm{LMO}(\mM^t) \in \argmin_{\mD \in \cS} \langle\mM^t, \mD\rangle\,,
\end{equation}
where $\mM^t$ is the effective update direction and $\cS \subset \Rmn$ is a constraint set. The algorithm proceeds as:
\begin{equation}\label{eq:simple_update}
  \begin{aligned}
  \mB^t &= \beta \mB^{t-1} + (1 - \beta) \nabla F(\mX^t) \quad \text{(momentum buffer)},\\
  \mM^t &= \begin{cases}
  \nabla F(\mX^t) & \text{(no momentum)},\\
  \mB^t & \text{(heavy ball)},\\
  \nabla F(\mX^t) + \beta \mB^t & \text{(approximate Nesterov)},
  \end{cases}\\
  \mX^{t + 1} &= \mX^t + \gamma_t\mathrm{LMO}\left(\mM^t\right)\,,
  \end{aligned}
\end{equation}
where $\beta \in [0, 1)$ is the momentum coefficient and $\nabla F(\mX^t)$ is either full or stochastic gradient. Throughout our experiments, we employ approximate Nesterov momentum, which matches the implementation in Muon and PyTorch SGD (\texttt{nesterov=True}). It uses the current gradient rather than the true lookahead gradient, trading theoretical guarantees for computational efficiency. From here on, we refer to stochastic gradient as just gradient if not otherwise specified.

@@

Using this formula, it is easy \citep{bernstein2024oldoptimizernewnorm, pethick2025training} to compute updates for algorithms induced by various norms $\norm{\cdot}$:

@@

\paragraph{Schatten norms}
\citep{cesista2025schattenp} considered Schatten-$p$ norms:
\[\norm{\mM^t}_{S_p} = \left(\sum_{i=1}^{\min(m, n)}\sigma_i^p\right)^{1/p}\,,
\]
which produce the update
\[
\mX^{t+1} = \mX^t - \gamma_t \mU \frac{\diag\left(\sigma_1^{q-1},\dots, \sigma_{\min(m,n)}^{q-1}\right)}{\left(\sum_{i=1}^{\min(m,n)}{\sigma_i^q}\right)^{\frac{q-1}{q}}}\mV^\top
\]
where $p$ and $q$ satisfy $p^{-1} + q^{-1} = 1$. This formula recovers Neon when $p\rightarrow1$ (provided that $\sigma_1 > \sigma_2$, which holds on real data), NSGD when $p=2$, and Muon when $p\rightarrow\infty$.

However, Schatten norms do not fill the rank gap: when $p > 1$, the update has full rank, while when $p=1$, its rank equals 1. Moreover, computing the update for $p \neq 1, 2, \infty$ appears to require knowing all $\sigma_i$, making the problem as hard as computing the full SVD.

@@

To derive the update formula for arbitrary $k$, we use the expression for the norm dual to the Ky Fan $k$-norm (see, for instance \citep{bhatia2013matrix}, p. 96):
\[
    \normkfk{\cdot}^\dagger = \max\left\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\right\}\,.
\]
According to \equationref{eq:our_update}, we need to find $\mD$ such that $\normkfk{\mD} = 1$ and
\[
\<\mM^t, \mD> = \max\left\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i,\ \sigma_1\right\}\,.
\]
The Neon update $\mD = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\mD = \frac{1}{k}\mU\mV^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
\[
\mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
\]
depending on which of those two expressions achieves smaller residual on the linear function $L(\mX^{t+1}) := F(\mX^t) + \<\mM^t, \mX^{t+1} - \mX^t>$. The Ky Fan norms thus fail to close the rank gap: the resulting update is either rank-one or full-rank.

@@

In \sectionref{sec:matrix-side}, we discuss how to efficiently compute Fanion updates using the Lanczos algorithm.

@@

%Simply applying norms dual to Ky Fan $k$-norms, however, does not provide sufficient algorithmic diversity for our purposes.
The approaches to design new LMO algorithms are not limited to applying norms dual to Ky Fan $k$-norms. We now consider linear combinations of LMO-based algorithms and show that these combinations are themselves LMO-algorithms.

@@

\begin{corollary}\label{corollary:lin_comb_alg}
	Let there be a finite family of LMO-based algorithms indexed by $i = 1,\dots,n$, where the $\mX^{t + 1}$ update in the $i$-th algorithm is defined by
	\[
	\mX^{t + 1} - \mX^{t} = \gamma_t \mathrm{LMO}_{i}(\mM^{t})\,,
	\]
	and $\mathrm{LMO}_i$ corresponds to the unit ball of norm $\norm{\cdot}_i$. For arbitrary non-negative $\alpha_1,\dots,\alpha_n$, the algorithm with the update given by
	\[
	\mX^{t + 1} - \mX^{t} = \gamma_t \sum_{i = 1}^n \alpha_i \mathrm{LMO}_{\norm{\cdot}_{(i)}}(\mM^t)
	\]
	is an LMO-algorithm itself, with LMO corresponding to the unit ball of the norm $\norm{\cdot}$ dual to the norm given by $\sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}^\dagger$.
\end{corollary}

@@

The extreme members of this family are \emph{F-Neon} (with $k = 1$) and \emph{F-Muon} (with $k = \min\{m, n\}$). Additional information and visualizations for the F-Muon generating $\normfstar{\cdot}^\dagger = \norm{\cdot}_{\mathrm{F-KF-1}}^\dagger$ and the F-Neon generating $\normftwo{\cdot}^\dagger = \norm{\cdot}_{\mathrm{F-KF-\min\{m, n\}}}^\dagger$ norms appear in the appendix (see \equationref{eq:normfstardual}, \equationref{eq:normftwodual}, \figureref{fig:fduals}).

@@

The extreme members of this family are \emph{S-Neon} (with $k = 1$) and \emph{S-Muon} (with $k = \min\{m, n\}$).

\section{Computing the Updates}\label{sec:matrix-side}
We employ the thick-restart the Lanczos method for the symmetric eigenvalue problem (TRLan) to compute the low-rank updates of Fanions. We apply TRLan to either $\mM^{t\top}\mM^t$ or $\mM^t \mM^{t\top}$, selecting whichever matrix is smaller. We use the CuPy implementation of \texttt{cupy.sparse.linalg.svds} \citep{cupy_svds_ref}, which internally relies on TRLan \citep{simonz1998thick}.

@@

\begin{table}[ht]
\floatconts
  {tbl:matrix_methods}
  {\caption{Comparison of methods for computing rank-$k$ updates on dense random matrices (CPU, double precision). Lower is better in all columns.}}
  {%
    \begin{tabular}{cccccccc}
      \arrayrulecolor{black}\toprule
      Matrix sizes & $k$  & Method           & Time (s) & Matvecs & Iterations & $err_1$         & $err_2$       \\
      \arrayrulecolor{black}\hline
      $500 \times 500$      & 5  & Power Iterations           & 0.062   & 2005    & 200        & $9.2 \cdot 10^{-3}$ & $9.1 \cdot 10^{-3}$\\
      $500 \times 500$      & 5  & RSVD             & 0.017   & 1170    & 38         & $9.8 \cdot 10^{-3}$ & $9.6 \cdot 10^{-3}$\\
      $500 \times 500$      & 5  & TRLan            & 0.018   & 131     & 65         & $9.6 \cdot 10^{-5}$ & $9.4 \cdot 10^{-5}$\\
      \hdashline
      $500 \times 500$      & 50 & Power Iterations           & 0.44    & 43750   & 437      & $9.9 \cdot 10^{-3}$ & $9.0 \cdot 10^{-3}$\\
      $500 \times 500$      & 50 & RSVD             & 0.61    & 6120    & 50      & $9.9 \cdot 10^{-3}$ & $9.1 \cdot 10^{-3}$\\
      $500 \times 500$      & 50 & TRLan            & 0.16    & 462     & 231      & $3.3 \cdot 10^{-7}$ & $3.0 \cdot 10^{-7}$\\
      \hdashline
      $5000 \times 5000$    & 5  & Power Iterations           & 9.6     & 9065    & 906      & $8.6 \cdot 10^{-3}$ & $8.6 \cdot 10^{-3}$\\
      $5000 \times 5000$    & 5  & RSVD             & 2.1     & 5640    & 187      & $9.7 \cdot 10^{-3}$ & $9.7 \cdot 10^{-3}$\\
      $5000 \times 5000$    & 5  & TRLan            & 0.70    & 205     & 102      & $7.7 \cdot 10^{-3}$ & $7.7 \cdot 10^{-3}$\\
      \arrayrulecolor{black}\bottomrule
    \end{tabular}%
  }
\end{table}

\begin{table}[ht]
\floatconts
  {tbl:newton_schulz_reference}
  {\caption{Newton-Schulz iterations performance on random dense matrices (for reference).}}
  {%
    \begin{tabular}{ccccc}
      \arrayrulecolor{black}\toprule
      Matrix size    & Time (s) & Matvecs & Iterations & $err_1$ \\
      \arrayrulecolor{black}\hline
      $500 \times 500$     & 0.041    & 27 000  & 27         & $4.8 \cdot 10^{-3}$  \\
      $5000 \times 5000$   & 26.4     & 290 000 & 29         & $6.5 \cdot 10^{-3}$  \\
      \arrayrulecolor{black}\bottomrule
    \end{tabular}%
  }
\end{table}

@@

We begin by evaluating F-Fanions and S-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ represent typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, and $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed on the interval $(0, 1)$. We set $\mS=0$ and initialize $\mX^0$ with entries drawn independently from $\cN(0, 0.01)$.

@@

\begin{table}[htbp]
\floatconts
  {tbl:lls_lrs}
  {\caption{Tuned learning rates and momentum coefficients in the linear least squares experiment.}}
  {%
    \begin{tabular}{cccc}
      \arrayrulecolor{black}\toprule
      Algorithm & \texttt{lr}  & \texttt{momentum}  & Iterations to 0.001 loss \\
      \arrayrulecolor{black}\hline
      Muon & 0.007 & 0.5 & 1060 \\
      \hdashline
      NSGD & 0.08 & 0.95 & 1020 \\
      F-Muon & 0.015 & 0.7 & 910 \\
      \hdashline
      SignSGD & 0.016 $\times$ 0.01 & 0.95 & 2650 \\
      S-Muon & 0.011 & 0.9 & 890 \\
      \arrayrulecolor{black}\bottomrule
    \end{tabular}%
  }
\end{table}

@@

\begin{figure}[htbp]
\floatconts
  {fig:lls}
  {\caption{Linear least squares problem for a 500x500 matrix.}}
  {%
    \subfigure[The loss]{\label{fig:lls_loss}%
      \includegraphics[width=0.8\linewidth]{figs/simple_lls/loss_long_run_16_increase.pdf}}%
    \vspace{1em}
    \subfigure[The Frobenius norm of the full gradient]{\label{fig:lls_fro_grad_norm}%
      \includegraphics[width=0.8\linewidth]{figs/simple_lls/gradient_frobenius_norm_vs_iteration_long_run.pdf}}
  }
\end{figure}

Both F-Muon and S-Muon converge faster to lower loss values and achieve lower Frobenius norms of the full gradient than NSGD, Muon, or SignSGD.

@@

\begin{table}[htbp]
\floatconts
  {tbl:mat_vs_vec_lmo}
  {\caption{LMO optimizers in Schatten $S_p$ norms and in $l_p$ norms. $g$ is the gradient. When it is a matrix, $g = \mU \mSigma \mV^\top$.}}
  {%
    \begingroup
    \def\arraystretch{1.2}
    \resizebox{\linewidth}{!}{%
      \begin{tabular}{lccc}
        \arrayrulecolor{black}\toprule
        Algorithm                             & LMO constraint set $\mathcal{D}$ & LMO                                                         & Reference                            \\
        \arrayrulecolor{black}\hline
        \arrayrulecolor{black}\hline
        Normalized SGD                     & $l_2$-ball, $S_2$-ball          & $-\eta \tfrac{g}{\norm{g}_2} = -\eta \tfrac{g}{\normf{g}}$ & \citep{hazan2015beyond}              \\
        Momentum Normalized SGD            & Ball in $l_2$, or Ball in $S_2$ & $-\eta \tfrac{g}{\norm{g}_2} = -\eta\tfrac{g}{\normf{g}}$  & \citep{cutkosky2020momentum}         \\
        \arrayrulecolor{black}\hline
        SignSGD                            & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 1]{bernstein2018signsgd} \\
        Signum                             & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 3]{bernstein2018signsgd} \\
        \hdashline
        Muon                               & Ball in Spectral $S_\infty$     & $-\eta \mU\mV^\top$                                             & \citep{jordan2024muon}               \\
        \arrayrulecolor{black}\hline
        Gauss-Southwell Coordinate Descent & Ball in $l_1$                   & $-\eta \sum_{i \in \argmax|g_i^t|} \sign(g_i^t)e_i$                       & \citep[p. 19]{shi2016primer}         \\
        \hdashline
        Neon                               & Ball in Nuclear $S_1$           & $-\eta \vu_1 \vv_1^\top$                                        & This work                            \\
        \arrayrulecolor{black}\bottomrule
      \end{tabular}%
    }%
    \endgroup
  }
\end{table}

@@

A common limitation of these analyses is their focus on convergence measured by the gradient norm. As we showed in our CIFAR experiments \sectionref{sec:cifar_underhood}, the stochastic gradient norm may decrease by only a factor of ten when the accuracy reaches 100\%.

@@

\subsection{The Nuclear Norm in the LMO}

@@

\subsection{Possible Improvements}

@@

\section{Visualization of Different Matrix Norms}
\subsection{Duals to F* and F2 Norms}

@@

\subsection{The Ky Fan Norm and Its Dual}

@@

\section{More Details for Linear Least Squares}
\label{sec:lls_plots_section}
The spectral and nuclear norms of the full gradients over iterations, as well as the loss over time, are shown in \figureref{fig:app_lls}. The poor performance of Fanions and F-Fanions in terms of speed for large $k$ is likely caused by the inefficient implementation of TRLan.
